Making Sequential Programs Deterministic

### 2019-04-06 design.tex
Edits in 2.6

\item
  Basetype polymorphism lets the basetype vary indepentently of the
  qualifier, which might be fixed or might be polymorphic.
  An example is a nondeterministic \<choose> function, which might have type

@@

\todo{Show actual type.}

@@

\subsection{Maps}\label{maps}
A map is a collection of key--value pairs.
Like other collections, a map or set can be nondeterministic,
order-nondeterministic, or deterministic.

@@

\subsection{Improving precision}\label{precision}

\todo{Sets vs lists.  Set equality is independent of order.}
An order nondeterministic list does not provide any guarantee on the
equality as it is dependent on the iteration order

We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("up") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \begin{align*}
    \infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det> \ \|boolean|}
    {
        &\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} \\
        & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List>        \hspace{150pt}   \\
        & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List>     \hspace{150pt}    
    }
    \end{align*}
    \bigskip

@@

\todo{
  In the JDK, the \<equals()> method of \<Set> is annotated as

@PolyDet("up") boolean equals(@PolyDet Set<E> this, @PolyDet Object o);

While this is sound, it is imprecise.
For example, consider the code below:

    @OrderNonDet Set<@Det Integer> a; \\
    @OrderNonDet Set<@Det Integer> b; \\
    a.equals(b);

Here, the return type of \<equals()> gets the annotation \<@NonDet>.
But it would be more precise to annotate it as \<@Det>.
To achieve this precision, the checker annotates the return type
of \<equals()> called on a receiver of type \<Set> as \<@Det> if the
following conditions are true:
\begin{itemize}
    \item the receiver does not have {@code List} or its subtype as a type parameter,
    \item the argument to equals() is also an {@code @OrderNonDet Set}, and
    \item the argument to equals() also does not have {@code List} or its subtype as a
    type parameter
\end{itemize}
}

According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List>} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List}.

Similarly, the rule \rulename{map precision}\ in \cref{fig-precision-rules}
improves precision for \<OrderNonDet> receivers.

% LocalWords:  basetypes NonDet OrderNonDet Det Iterable formedness param1
% LocalWords:  PolyDet boolean param2 arg1 arg2 localList entrySet keySet
% LocalWords:  LinkedHashMap TreeMap HashSet LinkedHashSet LinkedHashSet's

>>>>

\item
  Basetype polymorphism lets the basetype vary independently of the
  qualifier, which might be fixed or might be polymorphic.
  An example is a nondeterministic \<choose> function, which might have type

@@

@@

\subsection{Maps and sets}\label{maps}
A map is a collection of key--value pairs.
Like other collections, a map or set can be nondeterministic,
order-nondeterministic, or deterministic.

@@

\subsubsection{Improving precision for equality and map lookup}\label{precision}

List equality is dependent on iteration order, but set equality is not.
Comparing two objects of type \ptype{OrderNonDet List}{Det String} yields a \<NonDet>
result:  depending on the execution, the lists might or might not be in the
same order.  However, 
comparing two objects of type \ptype{OrderNonDet Set}{Det String} yields a
\<Det> result:  always the same on every execution.

\begin{figure}
    \begin{align*}
    \infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det> \ \|boolean|}
    {
        &\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} \hspace{10pt}\\
        & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> 
        \qquad \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List>     \hspace{30pt}
    }
    \end{align*}
    \bigskip

@@

More specifically, rule \rulename{set precision} of
\cref{fig-precision-rules} expresses that the return type of \<equals()>
is \<Det> if both arguments have type \<OrderNonDet Set> and nether argument
has {@code List} or a subtype within its type argument.
Without this rule, the type would be \<NonDet> which is sound but imprecise.

Similarly, the rule \rulename{map precision}\ in \cref{fig-precision-rules}
improves precision when calling \<get> on an \<OrderNonDet> receiver.

% LocalWords:  basetypes NonDet OrderNonDet Det Iterable formedness param1
% LocalWords:  PolyDet boolean param2 arg1 arg2 localList entrySet keySet
% LocalWords:  LinkedHashMap TreeMap HashSet LinkedHashSet LinkedHashSet's
% LocalWords:  basetype ddlist typechecked typecheck ni di
